Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-12-18 Manuscript/manuscript.tex
Update after comments by Kolafa

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD force fields has been considered a serious issue since the early days of lipid bilayer simulations. In this work, we circumvent the demanding explicit inclusion of electronic polarization effects~\cite{lucas12, chowdhary13} by accounting for the electronic part of polarizability in lipid bilayer simulations implicitly via the electronic continuum correction (ECC)~\cite{leontyev11}. Technically, this is similar to the phenomenological charge-scaling applied in earlier studies of surfactants, lipids or ionic liquids \cite{jonsson86, egberts94, beichel14}. However, the present concept of ECC is physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

@@

While using the scaling factor of $f_q = 0.75$ for ions in water is well justified in the ECC theory \cite{leontyev11}, it is not clear whether the same factor should be applied to partial charges used to describe molecules in MD models, e.g., lipids in our case. Unlike the total charge of an ion, atomic partial charges within molecules are not physical observables. Several schemes exist for the assignment of partial charges for biomolecules~\cite{Hu2007}, with the restrained electrostatic potential method (RESP) being commonly used~\cite{RESP_paper, Singh1984}. Considering that water is often included in RESP calculations or charges are refined to improve certain experimental observables, the electronic polarizability effects of the solvent may to some extent be included in standard force fields~\cite{RESP_paper, Singh1984, jorgensen96, ipolq2013, benavides17}. Thus, our application of the scaling factor, $f_q$, to existing partial charges in molecules does not necessarily follow $f_q = \epsilon _{el} ^{-1/2}$. Instead, a consistent scaling factor should lie between the value of 0.75  (\textit{i.e.,} no electronic polarizability included in the original partial charges) and 1 (\textit{i.e.,} electronic polarizability fully included in the original partial charges).

@@

To reproduce the experimental ion binding affinities,
we scanned possible values of the scaling factor from the interval $f_q~\in~(0.75, 1.0)$.
The ion binding affinity was benchmarked against experiments
using the head group order parameters and the electrometer concept \cite{seelig87,catte16},
as discussed more detail in the next section.
Scaling down the partial charges reduced the ion binding affinity.
We found the most accurate ion binding affinities with a scaling factor of $f_q = 0.8$,
which is only slightly higher than the ECC one ($f_q=0.75$).
Directly applying
the 0.8 scaling to the partial charges of the head group, the glycerol backbone, and
the carbonyls reduced the area per lipid to 60~\AA$^2$. This area is smaller than in the
original Lipid14 model ($65.6 \pm 0.5$~\AA$^2$)~\cite{dickson14} and in experiments
(64.3~\AA$^2$) \cite{kucerka11}. The decrease of the area per lipid arises from a
reduced hydration of the lipid head group region after scaling of the partial charges, which effectively
reduces the head group polarity. We solve this problem by slightly reducing the effective radii of
the modified head group atoms via lowering the $\sigma$ parameters in the Lennard-Jones potential by a
factor of $f_\sigma = 0.89$. The same approach was successfully adopted for the ECC ions in aqueous
solutions previously \cite{kohagen14, kohagen16, Pluharova2014, martinek17}. After reducing the head group atoms $\sigma$ parameters, the area per molecule is restored to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Comparing MD simulation to NMR experiments, we can validate the ion
binding affinity in lipid bilayer simulations using the 'electrometer concept'~ \cite{seelig87, catte16}.
This method is based on the experimental observation that the C-H bond order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group (Fig.~\ref{simVSexpNOions}) are proportional to the amount of charge bound per lipid~\cite{seelig87}.
The order parameters for all C-H bonds in lipid molecules can be accurately measured using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the order parameters can be calculated using the relation
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal. Angular brackets point to the average over all sampled configurations.

@@

To quantify the ion binding affinities to a lipid bilayer, we calculate the relative surface excess of ions with respect to water, $\Gamma_{ion}^{water}$~\cite{chattorajBOOK}. Such a quantity compares the adsorption of ions to the adsorption of water molecules at the interface without the necessity of defining a Gibbs dividing surface between the membrane interior and the water bulk region. In our simulations, we only assume that the interface is located between the ion-free hydrophobic interior of the lipid bilayer and the aqueous region far from the membrane.  Such a setup and the above definition of bulk ion concentrations provides a simple relation for the relative surface excess $\Gamma_{ion}^{water}$ for simulations of lipid bilayers,
 \begin{equation}\label{surfexcess}
  \Gamma_{ion}^{water}=\frac{1}{2A_b} \left ( n_{ion} - n_{water} \frac{C_{ion}}{C_{water}} \right ).
\end{equation}
Here, $n_{water}$ and $n_{ion}$ are the total numbers of water molecules and ions in the system,
$C_{water}$ and $C_{ion}$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

@@

First, we present results for bilayers in pure water.
The ECC-POPC and Lipid14 models both reproduce the experimental X-ray scattering form factors
of a POPC bilayer with a comparable accuracy (see Fig.~\ref{simVSexpNOions}).
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-POPC model
is $\approx$1\AA{} smaller than the experimental one.
The values of the area per lipid of the ECC-POPC model vary slightly
when simulated with different water models (i.e., within the interval of 62.2--66.8 \AA{}, see Table~S2 in SI),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-POPC model reproduces the experimental dimensions of the POPC
lipid bilayer with a comparable accuracy to other state-of-the-art lipid models~\cite{ollila16}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/isocontours_r37_ca_O-carb.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
    Isocontours of spatial number density of \ce{Ca^{2+}} (dark blue, 0.0016~\AA$^{-3}$) 
    and POPC carbonyl oxygen atoms (light semi-transparent red, 0.007~\AA$^{-3}$, all POPC lipids contribute). 
    Calcium cations localize mostly around phosphate oxygens (red and bronze).
    Interactions with carbonyl oxygens is less likely than with phosphate oxygens, 
    and it is contributed more by other neighbouring phospholipids than by the same lipid. 
    Transparent structures are shown to depict the variability of choline configurations 
    (colour warps from red to blue along the simulation time). 
    The number density was evaluated for each lipid, 
    after its strucutral alignment using only phosphate group.
    MDAnalysis \cite{mdanalysis2011} library was used for 
    the calculations of the structural alignment and the spatial number density. 
    VMD \cite{hump96} was used for visualisation. 
    Carbon atoms are cyan, hydrogen atoms white, oxygen atoms red, nitrogen blue.
  }
 %\todo{The figure should be updated as discussed last week. Joe: Will only smoothen the isosurfaces now. }
\end{figure}

@@

In this work, we reproduce the same experimental data used to infer binding stoichiometries employing our ECC-POPC model. Thanks to our simulations, we have a direct access to atomistic details of the binding stoichiometry without a need for any binding model as employed for interpreting in experiments~\cite{altenbach84, macdonald87}. To evaluate the relative propensities for each of the stoichiometric complexes (i.e.,~1~Ca$^{2+}$:~n~POPC), we calculated for each bound Ca$^{2+}$ the number of POPC molecules within a distance of 0.3~nm.
Results from the POPC bilayer simulation with a 285~mM bulk concentration of CaCl$_2$ are shown in Fig.~\ref{fig:cacl_complexes}.
We found the largest propensity for the 1:2 complex (41\%), with probabilities of complexes with the stoichiometries of 1:1~(25\%)~and 1:3~(34\%)~being only slightly lower. This suggests a more complex binding model than considered in a simple 1:2 ternary complex model previously. Nevertheless, with a broad brushstroke, the simulation data can be viewed such that one calcium binds to two lipids on average, because the probabilities of the complexes with 1 or 3 lipids are almost equal to each other  (and complexes with more than three lipids per one calcium ion were not observed). This probably explains why the simple the ternary complex model fits adequately the experimental data, as well as the ECC-POPC simulation results (see Fig.~S3 in SI).

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD simulation force fields has been considered a serious issue since the early days of lipid bilayer simulations. In this work, we circumvent the demanding explicit inclusion of electronic polarization effects~\cite{lucas12, chowdhary13} by including the electronic part of polarizability in lipid bilayer simulations implicitly via the electronic continuum correction (ECC)~\cite{leontyev11}. Technically, this is similar to the phenomenological charge-scaling applied in earlier studies of surfactants, lipids or ionic liquids \cite{jonsson86, egberts94, beichel14}. However, the present concept of ECC is physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

>>>>

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD force fields has been considered a serious issue since the early days of lipid bilayer simulations. In this work, we circumvent the demanding explicit inclusion of electronic polarization effects~\cite{lucas12, chowdhary13} by accounting for the electronic part of polarizability in lipid bilayer simulations implicitly via the electronic continuum correction (ECC)~\cite{leontyev11}. Technically, this is similar to the phenomenological charge-scaling applied in earlier studies of surfactants, lipids or ionic liquids \cite{jonsson86, egberts94, beichel14}. However, the present concept of ECC is physically well justified and has obtained theoretical support~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

@@

While using the scaling factor of $f_q = 0.75$ for ions in water is well justified in the ECC theory \cite{leontyev11}, it is not clear whether the same factor should be applied to partial charges used to describe molecules in MD models, e.g., lipids in our case \cite{leontyev14}. Unlike the total charge of an ion, atomic partial charges within molecules are not physical observables. Several computational schemes exist for the assignment of partial charges for biomolecules~\cite{Hu2007}, with the restrained electrostatic potential method (RESP) being commonly used~\cite{RESP_paper, Singh1984}. Considering that water is often included in RESP calculations or charges are refined to improve certain experimental observables, the electronic polarizability effects of the solvent may to some extent be included in standard force fields~\cite{RESP_paper, Singh1984, jorgensen96, ipolq2013, benavides17}. Thus, our application of the scaling factor, $f_q$, to existing partial charges in molecules does not necessarily follow $f_q = \epsilon _{el} ^{-1/2}$. Instead, a consistent scaling factor should lie between the value of 0.75  (\textit{i.e.,} no electronic polarizability included in the original partial charges) and 1 (\textit{i.e.,} electronic polarizability fully included in the original partial charges).

@@

To reproduce the experimental ion binding affinities,
we scanned possible values of the scaling factor from the interval $f_q~\in~(0.75, 1.0)$.
The ion binding affinity was benchmarked against experiments
using the head group order parameters and the electrometer concept \cite{seelig87,catte16},
as discussed more detail in the next section.
Scaling down the partial charges reduced the ion binding affinity.
We found the most accurate ion binding affinities with a scaling factor of $f_q = 0.8$,
which is only slightly higher than the ECC one ($f_q=0.75$).
Note that common scaling factors for monovalent ions in water are 0.8 or
larger \cite{benavides17,skinner14,nacleps}.  In contrast, modern force fields
for ionic liquids suggest values 0.6--0.65 which are lower than $\epsilon^{-1/2}_{el}$\cite{holm14}.
Directly applying
the 0.8 scaling to the partial charges of the head group, the glycerol backbone, and
the carbonyls reduced the area per lipid to 60~\AA$^2$. This area is smaller than in the
original Lipid14 model ($65.6 \pm 0.5$~\AA$^2$)~\cite{dickson14} and in experiments
(64.3~\AA$^2$) \cite{kucerka11}. The decrease of the area per lipid arises from a
reduced hydration of the lipid head group region after scaling of the partial charges, which effectively
reduces the head group polarity. We solve this problem by slightly reducing the effective radii of
the modified head group atoms via lowering the $\sigma$ parameters in the Lennard-Jones potential by a
factor of $f_\sigma = 0.89$. The same approach was successfully adopted for the ECC ions in aqueous
solutions previously \cite{kohagen14, kohagen16, Pluharova2014, martinek17}. After reducing the head group atoms $\sigma$ parameters, the area per molecule is restored to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Comparing MD simulation to NMR experiments, we can validate the ion
binding affinity in lipid bilayer simulations using the 'electrometer concept'~ \cite{seelig87, catte16}.
This method is based on the experimental observation that the C-H bond order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group (Fig.~\ref{simVSexpNOions}) are proportional to the amount of charge bound per lipid~\cite{seelig87}.
The order parameters for all C-H bonds in lipid molecules can be accurately measured using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the order parameters can be calculated using the relation
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal. Angular brackets point to the average over all sampled configurations.

@@

To quantify the ion binding affinities to a lipid bilayer, we calculate the relative surface excess of ions with respect to water, $\Gamma_{ion}^{water}$~\cite{chattorajBOOK}. Such a quantity compares the adsorption of ions to the adsorption of water molecules at the interface without the necessity of defining a Gibbs dividing surface between the membrane interior and the water bulk region. In our simulations, we only assume that the interface is located between the ion-free hydrophobic interior of the lipid bilayer and the aqueous region far from the membrane.  Such a setup and the above definition of bulk ion concentrations provides a simple relation for the relative surface excess $\Gamma_{ion}^{water}$ for simulations of lipid bilayers,
 \begin{equation}\label{surfexcess}
  \Gamma_{ion}^{water}=\frac{1}{2A_b} \left ( n_{ion} - n_{water} \frac{C_{ion}}{C_{water}} \right ).
\end{equation}
Here, $n_{water}$ and $n_{ion}$ are the total numbers of water molecules and ions in the system,
$C_{water}$ and $C_{ion}$ are their respective bulk concentrations in the aqueous phase
(measured from the density of farthest point between periodic bilayers),
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

@@

First, we present results for bilayers in pure water.
The ECC-POPC and Lipid14 models both reproduce the experimental X-ray scattering form factors
of a POPC bilayer with a comparable accuracy (see Fig.~\ref{simVSexpNOions}).
The area per lipid from the Lipid14 model is by $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-POPC model
is by $\approx$1\AA{} smaller than the experimental one.
The values of the area per lipid of the ECC-POPC model vary slightly
when simulated with different water models (i.e., within the interval of 62.2--66.8 \AA{}, see Table~S2 in SI),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-POPC model reproduces the experimental dimensions of the POPC
lipid bilayer with a comparable accuracy to other state-of-the-art lipid models~\cite{ollila16}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/isocontours_r37_ca_O-carb.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
    Isocontours of spatial number density of \ce{Ca^{2+}} (dark blue, 0.0016~\AA$^{-3}$) 
    and POPC carbonyl oxygen atoms (light semi-transparent red, 0.007~\AA$^{-3}$, all POPC lipids contribute). 
    Calcium cations localize mostly around phosphate oxygens (oxygens red, phosphate bronze).
    Interactions with carbonyl oxygens is less likely than with phosphate oxygens, 
    and it is contributed more by other neighbouring phospholipids than by the same lipid. 
    Transparent structures are shown to depict the variability of choline configurations 
    (colour warps from red to blue along the simulation time). 
    The number density was evaluated for each lipid, 
    after its strucutral alignment using only phosphate group.
    MDAnalysis \cite{mdanalysis2011} library was used for 
    the calculations of the structural alignment and the spatial number density. 
    VMD \cite{hump96} was used for visualisation. 
    Carbon atoms are cyan, hydrogen atoms white, oxygen atoms red, nitrogen blue.
  }
 %\todo{The figure should be updated as discussed last week. Joe: Will only smoothen the isosurfaces now. }
\end{figure}

@@

In this work, we reproduce the same experimental data used to infer binding stoichiometries employing our ECC-POPC model. Thanks to our simulations, we have a direct access to atomistic details of the binding stoichiometry without a need for any binding model as employed for interpreting in experiments~\cite{altenbach84, macdonald87}.
To evaluate the relative propensities for each of the stoichiometric complexes (i.e.,~1~Ca$^{2+}$:~n~POPC),
we calculated for each bound Ca$^{2+}$ the number of POPC molecules having oxygen atoms within a distance of 0.3~nm.
Results from the POPC bilayer simulation with a 285~mM bulk concentration of CaCl$_2$ are shown in Fig.~\ref{fig:cacl_complexes}.
We found the largest propensity for the 1:2 complex (41\%), with probabilities of complexes with the stoichiometries of 1:1~(25\%)~and 1:3~(34\%)~being only slightly lower. This suggests a more complex binding model than considered in a simple 1:2 ternary complex model previously. Nevertheless, with a broad brushstroke, the simulation data can be viewed such that one calcium binds to two lipids on average, because the probabilities of the complexes with 1 or 3 lipids are almost equal to each other  (and complexes with more than three lipids per one calcium ion were not observed). This probably explains why the simple the ternary complex model fits adequately the experimental data, as well as the ECC-POPC simulation results (see Fig.~S3 in SI).

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD simulation force fields has been considered a serious issue since the early days of lipid bilayer simulations. In this work, we circumvent the demanding explicit inclusion of electronic polarization effects~\cite{lucas12, chowdhary13} by including the electronic part of polarizability in lipid bilayer simulations implicitly via the electronic continuum correction (ECC)~\cite{leontyev11}. Technically, this is similar to the phenomenological charge-scaling applied in earlier studies of surfactants, lipids or ionic liquids \cite{jonsson86, egberts94, beichel14}. However, the present concept of ECC is physically well justified and has
obtained theoretical support~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.
